$p$-adic precision

### 2014-01-29 padicprec.tex
Differentials of usual operations on p-adic numbers and polynomials

\newcommand{\Z}{\mathbb Z}
\newcommand{\Zp}{\Z_p}
\newcommand{\Q}{\mathbb Q}
\newcommand{\Qp}{\Q_p}
\newcommand{\R}{\mathbb R}
\renewcommand{\O}{\mathcal O}

@@

%
% --- Author Metadata here ---
%\conferenceinfo{ISSAC}{}

% --- End of Author Metadata ---

\title{p-adic precision}

%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

%\numberofauthors{3} 
%\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
%\alignauthor
%Xavier Caruso\\
%       \affaddr{CNRS/Université de Rennes I}\\
%       \email{xavier.caruso@univ-rennes1.fr}
% 2nd. author
%\alignauthor
%David Roe\\
%       \affaddr{University of Calgary}\\
%       \email{roed.math@gmail.com}
% 3rd. author
%\alignauthor 
%Tristan Vaccon\\
%       \affaddr{Université de Rennes 1}\\
%       \email{tristan.vaccon@univ-rennes1.fr}
%}

\date{november 2013}

\maketitle
\begin{abstract}
A method to handle $p$-adic precision, to handle \textit{precision balance}.
\end{abstract}

% A category with the (minimum) three required fields
%\category{I.1.2}{Computing Methodologies}{Symbolic and Algebraic Manipulations}[Algebraic Algorithms]

%\terms{Algorithms, Theory}

%\keywords{$p$-adic precision, $p$-adic algorithm}

@@

Two approaches are possible. The first one consists in doing the 
computation step-by-step: we first compute $(z,t) = f(x,y)$ and we 
compute then $f(z,t)$ which gives the result. Proceeding this way, we 
see that $z$ and $t$ are both known up to precision $O(p^{\min(n,m)})$ 
are consequently the same holds for the coordinates of the result.

The second approach consists in remarking that $f \circ f$ maps
$(x,y)$ to $(2x, 2y)$. CComputing $f \circ f(x,y)$ using the latter
formula, it is clear that the result we get is known up to precision
$(O(p^n), O(p^m))$ --- and even $(O(p^{n+1}), O(p^{m+1}))$ if $p$
equals $2$.

As a conclusion, the second approach yields a more accurate result.
In other words, computing step-by-step creates ``artificial'' losses 
of precision.

@@

If $M$ is random square matrix over $\Zp$ of side $d$ whose every
coefficient is known up to precision $O(p^N)$, the two following
results are proved in \emph{loc. cit.}:
\begin{itemize}
\item using usual Gauss elimination and tracking $p$-adic precision 
step-by-step, the smallest precision on an entry of $L(M)$ is about
$O(p^{N - \frac{2d}{p-1}})$ on average;
\item computing $L(M)$ by evaluating Cramer-type formulas yields a
result whose every entry is known up to precision $O(p^{N - 2 \log_p 
d})$.
\end{itemize}
If $d$ is large compared to $p$, the second precision is much more
accurate than the first one. This shows once again that ``artifical''
losses of precision appear when one performs Gauss elimination by
tracking precision step-by-step.

@@

\todofor{Xavier}{Write this section}

\subsection{Differential of usual operations}

\subsubsection{Numbers}

\todo{sum, product, inverse, $p$-power map}

\subsubsection{Polynomials}

\todo{
Euclidean division, factorization, Grobner basis (?)
}

\subsubsection{Matrices}

\todo{
determinant, characteristic polynomial, LU factorization, QR 
factorization, kernels, images}

\subsubsection{Sub vector spaces}

\todo{intersection, sum}

@@

\section{Conclusions}

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
%APPENDICES are optional
%\balancecolumns

\tiny \bibliographystyle{plain}
\begin{thebibliography}{10}
\addcontentsline{toc}{section}{Références}

@@

\bibitem{Schneider}
  P.~Schneider,
  \emph{$p$-adic Lie Groups}, Springer (2011)
\end{thebibliography}

>>>>

\newcommand{\Z}{\mathbb Z}
\newcommand{\Zp}{\Z_p}
\newcommand{\Q}{\mathbb Q}
\newcommand{\Qp}{\Q_p}
\newcommand{\R}{\mathbb R}
\renewcommand{\O}{\mathcal O}
\newcommand{\XX}{\mathbf X}

@@

\title{$p$-adic precision}
\author{Xavier Caruso}
\author{David Roe}
\author{Tristan Vaccon}
\date\today

\maketitle
\begin{abstract}
We present a new method to handle $p$-adic precision (or more generally
precision over ultrametric fields) based on $p$-adic calculus.
We illustrate it by many examples and give a toy application to the
stable computation of the SOMOS 4 sequence.
\end{abstract}

\setcounter{tocdepth}{1}
\tableofcontents

@@

Two approaches are possible. The first one consists in doing the 
computation step-by-step: we first compute $(z,t) = f(x,y)$ and we 
compute then $f(z,t)$ which gives the result. Proceeding this way, we 
see that $z$ and $t$ are both known up to precision $O(p^{\min(n,m)})$ 
are consequently the same holds for the coordinates of the result.
The second approach consists in remarking that $f \circ f$ maps
$(x,y)$ to $(2x, 2y)$. Computing $f \circ f(x,y)$ using the latter
formula, it is clear that the result we get is known up to precision
$(O(p^n), O(p^m))$ --- and even $(O(p^{n+1}), O(p^{m+1}))$ if $p$
equals $2$.
As a conclusion, the second approach yields a more accurate result.
In other words, computing step-by-step creates ``artificial'' losses 
of precision.

@@

If $M$ is random square matrix over $\Zp$ of side $d$ whose every
coefficient is known up to precision $O(p^N)$, the two following
results are proved in \cite{Caruso}:
\begin{itemize}
\item using usual Gauss elimination and tracking $p$-adic precision 
step-by-step, the smallest precision on an entry of $L(M)$ is about
$O(p^{N - \frac{2d}{p-1}})$ on average;
\item computing $L(M)$ by evaluating Cramer-type formulas yields a
result whose every entry is known up to precision $O(p^{N - 2 \log_p 
d})$.
\end{itemize}
If $d$ is large compared to $p$, the second precision is much more
accurate than the first one. This shows once again that ``artifical''
losses of precision appear when one performs Gauss elimination by
tracking precision step-by-step.

@@

In this Section, we illustrate the theory developed in this paper by 
giving examples and applications. In \S \ref{ssec:differentials}, we 
compute the differential of a bunch of usual operations involving 
$p$-adic objects of different nature. Very often, we will observe that 
this computation is easy even if the underlying operation is 
sophisticated (\emph{e.g.} Gr\"obner basis).
In \S \ref{ssec:SOMOS-solution}, we study the SOMOS 4 sequence 
introduced in \S \ref{ssec:stepbystep} and, making a crucial use of 
Lemma \ref{lem:main} and Proposition \ref{prop:}, we design a quite 
\emph{stable} algorithm for computing it.

\subsection{Differential of usual operations}
\label{ssec:differentials}

As in the previous Sections, we always work over a fixed ultrametric 
field $K$, whose absolute value is denoted by $|\cdot|$. We recall that 
usual examples are the field of $p$-adic numbers $\Q_p$ and the field of 
Laurent series $k((X))$ over a base field $k$.

In what follows, we use freely the ``method of physicists'' to compute 
differentials: given a function $f$ differentiable at some point $x$, we 
consider a small perturbation $dx$ of $x$ and write $f(x+dx) = y + dy$ 
by expanding LHS and neglecting terms of order $2$. The differential of 
$f$ at $x$ is then the linear mapping $dx \mapsto dy$.

\subsubsection*{Numbers}

The most basic operations are, of course, sums, products and inverses of 
elements of $K$. Their differential are well-known and quite easy to 
compute: if $z = x + y$ (resp. $z = xy$, resp $z = \frac 1 x$), we have 
$dz = dx + dy$ (resp. $dz = x \cdot dy + dx \cdot y$, resp $dz = - 
\frac{dx}{x^2}$).

A little bit more interesting is the $n$-th power map $f$ from $K$ to 
itself. Its differential $f'(x)$ is obtained by differentiate the 
equality $y = x^n$: we obtain $dy = n x^{n-1} dx$.
Hence $f'(x)$ maps to ball $B_K(r)$ (where $r$ is some positive real 
number) to $B_K(|n|{\cdot} r)$. According to Lemma \ref{lem:main}, this 
means that the behaviour of the precision depends on the absolute value 
of $n$. By the ultrametric inegality, we always have $|n| \leq 1$ but 
this inegality might be strict. It happens for instance if the 
characteristic of $K$ is divisible by $p$, in which case $n$ vanishes in 
$K$. In that case $f'(x)$ also vanishes and Lemma \ref{lem:main} does 
not apply. 
Another situation (which is more interesting) is that of $p$-adic 
numbers: let us take $K = \Q_p$ and $n = p^k$ for some integer $k$. Then
$|n| = p^{-k}$ and Lemma \ref{lem:main} tells us basically that raising
a $p$-adic number to the $p^k$-th power increases the precision by $k$
extra digits (which is actually a well-known fact).

\subsubsection*{Univariate polynomials}

Beyond sums and products (which can be treated as before), a basic 
operation involving polynomials is Euclidean division. Let us compute 
the differential of it. We start with two polynomials $A$ and $B$ with 
$B \neq 0$. The Euclidean division of $A$ by $B$ is written $A = BQ + R$ 
with $\deg R < \deg B$. Differentiating the above equality we find: 
$dA - dB \cdot Q = B \cdot dQ + dR$,
which implies that $dQ$ and $dR$ are respectively obtained as the 
quotient and the remainder of the Euclidean division of $dA - dB \cdot 
Q$ by $B$. This gives the differential. We note that the discussion 
above extends readily to convergent series (see also 
\cite{caruso-lubicz}).

Another usual operations on polynomials is the factorization. Suppose 
that we are given a monic polynomial $P \in K[X]$ written as a product 
$P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. Hensel's 
lemma implies that there exists a small neighbourhood $\mathcal U$ of 
$P_0$ such that any monic polynomial $P \in \mathcal U$ factors uniquely 
as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ and $B_0$ 
respectively. Thus, we can consider the application $f : P \mapsto (A,B)$ 
defined on the subset of $\mathcal U$ consisting of monic polynomials.
We want to differentiate $f$ at $P_0$. For this, we differentiate the 
equality $P = A B$ around $P_0$, obtaining 
\begin{equation}
\label{eq:difffactor}
dP = A_0 \cdot dB + B_0 \cdot dA.
\end{equation}
where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
between $A_0$ and $B_0$, it follows from Eq.~\eqref{eq:difffactor} that
$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).

An important special case of the previous study occurs when $A_0$ has 
degree $1$, that is $A_0(X) = X - \alpha$ with some $\alpha \in K$. The 
application $P \mapsto A$ is then nothing but the mapping that follows 
the simple root $\alpha$. Of course, its differential around $P_0$ can 
be computed by the above method. Nevertheless there is a more direct way 
to obtain it in this particular case: we write $(P + dP)(\alpha + 
d\alpha) = 0$ and expand this. We obtain this way $P'(\alpha) d\alpha + 
dP(\alpha) = 0$ where $P'$ denotes the derivative of $P$. Since $\alpha$ 
is a simple root, $P'(\alpha)$ does not vanish and we recover $d \alpha 
= - \frac{dP(\alpha)}{P'(\alpha)}$.

\subsubsection*{Multivariate polynomials}

We consider the ring $K[\XX] = K[X_1, \ldots, X_n]$ of polynomials in $n$
variables over $K$ and fix a monomial order on it. We then have a 
notion of division in $K[\XX]$: if $f, f_1, \ldots, f_s$ are polynomials
in $K[\XX]$, there exists a writing:
$$f = q_1 f_1 + \cdots + q_s f_s + r$$
where $q_1, \ldots, q_s, r \in K[\XX]$ and no term of $r$ is divisible
by the leading term of some $f_i$ ($1 \leq i \leq s$). Moreover, assuming
that $(f_1, \ldots, f_s)$ is a Gr\"obner basis\footnote{Without this
assumption, there still exists a canonical choice of $r$. We do not know
if the computation of the differential that follows extends to a more
general setting.} (of the ideal generated by these polynomials), the 
polynomial $r$ is uniquely determined and called the \emph{remainder} of 
the division of $f$ by the family $(f_1, \ldots, f_s)$. The application 
$(f, f_1, \ldots, f_s) \mapsto r$ is then well defined and similarly to 
the case of univariate polynomials, we can compute its differential: we 
find that $dr$ is obtained as the remainder of the division of $f - (q_1 
\cdot d f_1 + \cdots + q_s \cdot d f_s)$ by $(f_1, \ldots, f_s)$.

We now focus on the computation of Gr\"obner basis. Recall that any 
ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We 
can then consider the application mapping a family $(f_1, \ldots, f_s)$ 
of polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
\ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
application at $(f_1, \ldots, f_s)$ is continuous provided that:
\begin{itemize}
\item the sequence $(f_1, \ldots, f_s)$ is regular
\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
$f_1, \ldots, f_i$ is wearly-$w$ where $w$ denotes the fixed monomial
order (\emph{cf} \cite{Vaccon} for more precisions).
\end{itemize}
A similar argument proves that it is actually differentiable at such
points. We are now going to compute the differential. For this we
remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
g_t)$ generate the same ideal, we have a relation:
\begin{equation}
\label{eq:prodgrobner}
(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
\end{equation}
where $A$ is a $(s \times t)$ matrix with coefficients in $K[\XX]$.
Moreover following Vaccon's construction, we see that the entries of $A$ 
can be chosen in such a way that they define differentiable functions.
Differentiating Eq.~\eqref{eq:prodgrobner}, we get:
$$(d f_1, \ldots, d f_s) \cdot A = - (f_1, \ldots, f_s) \cdot dA + 
(d g_1, \ldots, d g_t).$$
Finally, using that we are working with \emph{reduced} Gr\"obner
basis, the above equality implies that $d f_i$ is the remainder in
the division of the $i$-th coefficient of the product
$(d f_1, \ldots, d f_s) \cdot A$ by the family $(f_1, \ldots, f_s)$.

\subsubsection*{Matrices}

\todo{Determinant, characteristic polynomial, LU factorization, QR 
factorization, kernels, images}

\subsubsection*{Vector spaces}

\todo{Intersection, sum}

@@

%\section{Conclusions}

\begin{thebibliography}{10}

@@

\bibitem{Schneider}
  P.~Schneider,
  \emph{$p$-adic Lie Groups}, Springer (2011)

\bibitem{Vaccon}
  T.~Vaccon,
  \emph{Matrix-F5 algorithms over finite-precision complete discrete 
  valuation fields},
  preprint (2014)
\end{thebibliography}
